Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-12 aistats.tex
Reference a review of SGD assumptions
Write about Dragomir's assumption

%\hypersetup{draft}

@@

\subsection{Bounding the Randomness}
Smoothness is not the only issue.
To analyze stochastic algorithms, one also need to quantify the randomness. 
We now review these modern attempts at analyzing SMD, and illustrate their limits  in the context of the MAP.
See \cref{tbl:assumptions} for a summary.

\begin{table*}[t]\begingroup
\renewcommand*{\arraystretch}{1.25}%
\centering
\begin{tabular}{p{.35\textwidth}p{.6\textwidth}}
\toprule
Assumption &  \\
\midrule
Strongly convex and bounded variance\newline
\citep[e.g.,][]{??}
&
$
\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
\quad
\expect[x]{\norm{\m - x}}^2\leq\sigma^2
$
\\
Uniformly bounded ``variance''\newline
\citep{hanzely2018fastest}
&
	$
	\expect[x]{
	\begin{aligned}
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}\\
	+\, \cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\end{aligned}
	}
	\leq
	\eta C
	$
\\
Bounded ``variance'' at $\theta_*$\newline
\citep{dragomir2021fast}
&
	$\expect[x]{
		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
	} \leq 2 \eta^2 C$
\\
Bounded optimality gap\newline
\citep{dorazio2021stochastic}
&
?
\\
\bottomrule
\end{tabular}
\caption{Caption}
\label{tbl:assumptions}
\endgroup\end{table*}

\subsubsection{Bounded variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step
\alignn{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C, \forall n \geq 1
\label{eq:assumption-hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

@@

In general however, this assumption hardly holds.
For the MAP, this translates into
\alignn{
\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C, \forall n \geq 1
}
which fails to hold on the gaussian variance example for instance. \TODO{short proof or formalize.}

\subsubsection{Bounded variance at the minimum}

\citep{dragomir2021fast} is similar to \cref{eq:hanzely-in-divergence},
where we take only the first part of it
and care only about gradients at the minimum, replacing
$\tilde g(\theta)$ and  $\nabla f(\theta)$ with $\tilde g(\theta_*)$ and $\nabla f(\theta_*) = 0$
\aligns{
	\expect[\tilde g]{
		\cB_{A^*}(\m - 2\eta \tilde g(\theta_*), \m)
	} \leq 2 \eta^2 C.
}
{\bf For a maximum likelihood problem, }
\aligns{
	\expect[x]{
		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
	} \leq 2 \eta^2 C.
}
\fdk{How do we show that this fails?}

\subsubsection{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}
\aligns{
	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
}

@@

>>>>

\hypersetup{draft}

@@

\subsection{Bounding the Randomness}
Smoothness is not the only issue.
To analyze stochastic algorithms, one also need to quantify the randomness.
Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
See \cref{tbl:assumptions} for a summary.

Caveat : all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.

\begin{table*}[t]\begingroup
\renewcommand*{\arraystretch}{1.25}%
\centering
\begin{tabular}{p{.35\textwidth}p{.6\textwidth}}
\toprule
Assumption &  \\
\midrule
Strongly convex and bounded variance\newline
\citep[e.g.,][]{??}
&
$
\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
\quad
\expect[x]{\norm{\m - x}}^2\leq\sigma^2
$
\\
Uniformly bounded ``variance''\newline
\citep{hanzely2018fastest}
&
	$
	\expect[x]{
	\begin{aligned}
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}\\
	+\, \cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\end{aligned}
	}
	\leq
	\eta^2 C
	$
\\
Bounded ``variance'' at $\theta_*$\newline
\citep{dragomir2021fast}
&
	$\expect[x]{
		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
	} \leq 2 \eta^2 C$
\\
Bounded optimality gap\newline
\citep{dorazio2021stochastic}
&
?
\\
\bottomrule
\end{tabular}
\caption{Caption}
\label{tbl:assumptions}
\endgroup\end{table*}

\subsubsection{Bounded Variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
\alignn{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
\label{eq:assumption-hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

@@

In general however, this assumption hardly holds.
For the MAP, this translates into $\forall n \geq 1,$
\alignn{
\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C
}
which fails to hold on the gaussian variance example for instance. \TODO{short proof or formalize.}

\subsubsection{Bounded Variance at the Minimum}

\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; .
	\label{eq:assumption-dragomir}
}
In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
Under this assumption, they prove convergence up to a variance ball with a constant step-size.
Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.

Seems like a perfect fit then ? Well not really, this variance assumption does not hold for exactly the same reason as the one of Hanzely. Having something valid for every possible iterates is way too strong.

\subsubsection{Bounded Optimality Gap}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}
\aligns{
	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
}

@@

If assumptions~\eqref{eq:assumption-hanzely} and~\eqref{eq:assumption-dragomir} do not hold for every iterate, then let's assume they hold on average. Does the proof remain valid ? Yes I think so \TODO{check this}. Does it become impossible to prove / as hard as the initial problem ? Yes.
\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
